### 2024-09-26 AISTAT25/aistats.tex
Cleanup

\iffalse
\subsection{Results}

%Our algorithm does not require any assumption on the input distribution. In the worst case the number of queries is twice the number of queries of a passive algorithm. We show that under several natural conditions the query complexity of our algorithm is logarithmic in the error.

@@

The rest of the paper is organized as follows. xxxxxxx

@@

\section{A general-purpose active learning algorithm}
Our algorithm has several intricacies that might make it hard to
follow at a first read.  We therefor start with a description of a
simplified version of the algorithm (Fig~\ref{alg:siple}) that is not
completely correct but serves to explain the main components of the
detailed algorithm. The full description of the algorithm is given in
subsection~\ref{sec:full description} and appendix~\ref{sec:detailedalgorithm}

@@

In step ~\ref{step:balls} associates a label with each ball. The
label is one of $\{-1,+1,0\}$.  The labels $+1$ and $-1$ correspond to
a confident statement that $\eta(B)>0$ and $\eta(B)<0$
respectively. Finally, the label $0$ indicates that the sign of $B$
is uncertain given the answers to the queries.

@@

These bias estimates will with high probability be seen to satisfy the following guarantee.
\begin{defn}
For any $B \in \B$, bias estimate $\yh(B) \in \{+1,-1,0\}$ is \emph{$\gamma$-accurate} if:
\begin{itemize}
\item $\yh(B) = +1 \implies \eta_X(B) > 0$
\item $\yh(B) = -1 \implies \eta_X(B) < 0$
\item $\yh(B) = 0 \implies |\eta_X(B)| < \gamma$
\end{itemize}
\label{def:accurate-bias-estimate}
\end{defn}

\subsection{Assigning labels to points}

@@

Pick any $x \in X$ and any level $\ell$. Once qualitative bias estimates $\yh(B)$ are available for all balls at levels $\leq \ell$ that contain $x$ (that is, $\B_{\leq \ell}(x)$), we use these to infer a level-$\ell$ label for $x$. To begin with, the set of \emph{possible} labels for $x$ at level $\ell$, denoted $\PL_\ell(x) \subset \{-1,+1\}$, is defined thus:
\begin{itemize}[leftmargin=0.5cm]
\item $\PL_\ell(x)$ includes $+1$ if there exists a \emph{minimal} ball $B \in \B_{\leq \ell}(x)$ (with no other $B' \in \B_{\leq \ell}(x)$ strictly contained in it) for which $\yh(B) = +1$.
\item $\PL_\ell(x)$ includes $-1$ under a symmetrical condition.
\end{itemize}
This is made precise in Eqn~(\ref{eq:PL}) in Fig.~\ref{alg:main}. The label-estimate for $x$ at level $\ell$, denoted $\yh_\ell(x)$, is $+1$ if $\PL_\ell(x)=\{+1\}$, $-1$ if $\PL_\ell(x) = \{-1\}$, $0$ if $\PL_\ell(x) = \{\}$, and $!$ if $\PL_\ell(x) = \{-1,+1\}$ (Eqn~(\ref{eq:provisional-label})). 
%The interpretation of $\yh_\ell(x) \in \{+1,-1,0\}$ is given in Definition~\ref{def:accurate-bias-estimate}.

@@

\section{Detailed algorithm}

@@

Repeat:
\begin{itemize}[leftmargin=0.5cm]
\item If there is a level $\ell \geq 0$ such that $U_\ell \neq \emptyset$:
\begin{itemize}[leftmargin=0.25cm]
\item For the smallest such level $\ell'$, run {\tt Focused-query}($\ell', U_{\ell'}$) // Fig.~\ref{alg:sampling}
\end{itemize}
\item Run {\tt Background-query} // Fig.~\ref{alg:sampling}
\item Update labels:
\begin{itemize}[leftmargin=0.25cm]
\item Update bias-estimates $\yh(B)$ // Fig.~\ref{alg:bias-estimate}
\item For each $x \in X$ and level $\ell \geq 0$ for which all $\{\yh(B): B \in \B_\ell(x)\}$ are available:
\begin{align}
  \PL_\ell(x) &= \{s \in \{-1, +1\}: \\
             & \mbox{there exists $B \in \B_{\leq \ell}(x)$ s.t. $\yh(B) = s$ and} \notag \\ 
             & \mbox{no $B' \in \B_{\leq \ell}(x)$ has $X_{B'} \subsetneq X_B$.} \} \label{eq:PL} \\ 
\yh_\ell(x) &= 
\left\{
\begin{array}{cl}
+1 & \mbox{if $\PL_\ell(x) = \{+1\}$} \\
-1 &  \mbox{if $\PL_\ell(x) = \{-1\}$} \\
0 & \mbox{if $\PL_\ell(x) = \{\}$} \\
! & \mbox{if $\PL_\ell(x) = \{-1,+1\}$}
\end{array}
\right.
\label{eq:provisional-label}
\end{align}

@@

In Fig.~\ref{alg:sampling}, we see that the points queried in a ball $B$ at level $\ell$ are $\{z \in X_B: T_z \leq \tau_\ell\}$, which from the definition of $\tau_\ell$ has size $O(k)$. We use these labels to estimate the bias of $B$. Since we need to detect biases of magnitude greater than $\gamma$, we set $k \propto 1/\gamma^2$. 
%This intuition is borne out by the following result, proved in the appendix. 
\begin{lemma}
Suppose that $k \geq (192/\gamma^2) \ln (4 |\B|/\delta)$. 
Let $\yh(B)$ be defined as in Fig.~\ref{alg:bias-estimate}. Then with probability $\geq 1-\delta$, all the $\yh(B)$, for $B \in \B$, are $\gamma$-accurate in the sense of Definition~\ref{def:accurate-bias-estimate}.
\label{lemma:accurate-bias-estimates}
\end{lemma}
Henceforth assume $\yh(B)$ is $\gamma$-accurate for all  $B \in \B$.

@@

Recall that $\PL_\ell(x) \subset \{-1,+1\}$ denotes the possible labels for $x$ given information from balls at levels $\leq \ell$, and is used to assign a value $\yh_\ell(x) \in \{-1,+1,0,!\}$. Despite randomness in querying, it is possible to define two critical levels for each $x$: a level $L_1(x)$ by which $\PL_\ell(x)$ reliably contains the correct label of $x$, and a level $L_2(x)$ by which $\PL_\ell(x)$ reliably omits the wrong label.

@@

\begin{thm}[Local query complexity]
Take $k \geq (192/\gamma^2) \ln (4 |\B|/\delta)$. With probability at least $1-2\delta$, the following holds simultaneously for all $x \in X$. Let $L_1(x)$ and $L_2(x)$ be the critical levels for $x$, as in Definition~\ref{defn:L12}. If $L_2(x) \leq \lg (n/2k)$, let
\[
m_o(x) = 32k \cdot \max\bigg( 2^{L_1(x)}, \ \frac{1}{n} \sum_{\ell=L_1(x)+1}^{L_2(x)} |\Delta_\ell| \, 2^\ell\bigg) .
\]
Once the active learner has made $m_o(x)$ queries, $\yh(x)$ is fixed at the Bayes-optimal label $g^*(x)$.
\label{thm:label-complexity}
\end{thm}
Let's interpret this bound. For a specific $x \in X$, there is no telling how $\yh_\ell(x)$ will behave at levels $\ell < L_1(x)$. But for $\ell \geq L_1(x)$, either $\yh_\ell(x) = \, !$, in which case $x$ is in the uncertainty region, or the label of $x$ is correctly set once and for all, which occurs by level $L_2(x)$.

@@

\begin{figure}[t]
\begin{center}
  \includegraphics[width=2.5in]{figures/Monotone.png}
%\hspace{.5in}
\includegraphics[width=2.5in]{figures/Massart.png}
\end{center}
\caption{(a) The conditional probability function $\eta$ is monotonically increasing; $\eta(x) = -\gamma, 0, \gamma$ at $x = \lambda_L, \lambda, \lambda_R$, respectively. (b) On each subinterval, $\eta$ is either $> \gamma$ or $<-\gamma$.}
\label{fig:oned}
\end{figure}

@@

\subsection{Example: Massart noise}

In Section~\ref{sec:oned-massart}, we develop another one-dimensional example, in which $\X$ consists of several pieces, such that on each piece $\eta$ is either entirely above $\gamma$ or entirely below $-\gamma$ (Fig.~\ref{fig:oned}(b)). Once again, a logarithmic label complexity is obtained.

@@

We describe an experiment designed to evaluate our algorithm.
We use a simple 1D problem, described in figure~\ref{fig:oned}(b)
(Massard Noise).  Although this problem is simple, it is outside the scope of most non-parametric active learning algorithms, which assume smoothness of the conditional probability.

We compare two algorithms, the "active" agorithm described here, and the "passive" algorithm which uses only "background queries". We denote the difference between the error of the generated classifier and that of the Bayes optimal classifier by $\epsilon$. From the theory, we expect the query complexity of the passive algorithm to be $O(\frac{1}\epsilon})$ and of the active algorithm to be $O(\log (\frac{1}{\epsilon})$.

@@

\subsubsection{Accuracy of bias estimates: Proof of Lemma~\ref{lemma:accurate-bias-estimates}}

@@

For part (a), note that some $B_o \in \B_{L_1(x)}(x)$ has significant bias (that is, bias $\geq \gamma$) towards the correct label $s(x)$, as does any ball $B \in \B(x)$ contained within it. For $\ell \geq L_1(x)$, the set $\{B \in \B_{\leq \ell}(x): X_{B} \subseteq X_{B_o}\}$ is nonempty (it contains $B_o$); pick any minimal ball $B$ within it. Then $\yh(B) = s(x)$ by the $\gamma$-accuracy of bias estimates (Lemma~\ref{lemma:accurate-bias-estimates}) and thus $s(x) \in \PL_\ell(x)$.

@@

%\begin{figure}[t]
%\begin{center}
%\includegraphics[width=3in]{curvature.pdf}
%\end{center}
%\caption{{\bf A 2D example with bounded curvature:} {\em Left:} We have $\eta = -0.2$ for the orange/red part of the domain and $\eta=0.2$ for the light and dark blue. The orange and light blue indicate the boundary sets for negative and positive respectively. {\em Right:} examples of the $L_1,L_2$ balls for different points. The green $L_1$ circles are the largest monochrome circle that contains the point. The purple $L_2$ circles are the union of all circles of half the radius or smaller. All of these circles are monochrome.}
%\label{fig:curvature}
%\end{figure}

>>>>

@@

@@

\section{A general-purpose active learning algorithm}
\label{sec:algorithm}
Our algorithm has several intricacies that might make it hard to
follow at a first read.  We therefor start with a description of a
simplified version of the algorithm (Fig~\ref{alg:simple}) that is not
completely correct but serves to explain the main components of the
detailed algorithm. The detailed algorithm is described in
subsection~\ref{sec:detailedalgorithm}.

@@

Step ~\ref{step:balls} associates a label with each ball. The
label is one of $\{-1,+1,0\}$.  The labels $+1$ and $-1$ correspond to
a confident statement that $\eta(B)>0$ and $\eta(B)<0$
respectively. Finally, the label $0$ indicates that the sign of $B$
is uncertain given the answers to the queries.

@@

These bias estimates will with high probability be seen to satisfy the following guarantee.
\subsection{Assigning labels to points}

@@

Pick any $x \in X$ and any level $\ell$. Once qualitative bias estimates $\yh(B)$ are available for all balls at levels $\leq \ell$ that contain $x$ (that is, $\B_{\leq \ell}(x)$), we use these to infer a level-$\ell$ label for $x$. To begin with, the set of \emph{possible} labels for $x$ at level $\ell$, denoted $\PL_\ell(x) \subset \{-1,+1\}$, is defined thus:
\begin{itemize}[leftmargin=0.5cm]
\item $\PL_\ell(x)$ includes $+1$ if there exists a \emph{minimal} ball $B \in \B_{\leq \ell}(x)$ (with no other $B' \in \B_{\leq \ell}(x)$ strictly contained in it) for which $\yh(B) = +1$.
\item $\PL_\ell(x)$ includes $-1$ under a symmetrical condition.
\end{itemize}
This is made precise in Eqn~(\ref{eq:PL}) in Fig.~\ref{alg:main}. The label-estimate for $x$ at level $\ell$, denoted $\yh_\ell(x)$, is $+1$ if $\PL_\ell(x)=\{+1\}$, $-1$ if $\PL_\ell(x) = \{-1\}$, $0$ if $\PL_\ell(x) = \{\}$, and $!$ if $\PL_\ell(x) = \{-1,+1\}$ (Eqn~(\ref{eq:provisional-label})). 
The interpretation of $\yh_\ell(x) \in \{+1,-1,0\}$ is given in Definition~\ref{def:accurate-bias-estimate}.

@@

@@

Repeat:
\begin{itemize}[leftmargin=0.5cm]
\item If there is a level $\ell \geq 0$ such that $U_\ell \neq \emptyset$:
\begin{itemize}[leftmargin=0.25cm]
\item For the smallest such level $\ell'$, run {\tt Focused-query}($\ell', U_{\ell'}$) // Fig.~\ref{alg:sampling}
\end{itemize}
\item Run {\tt Background-query} // Fig.~\ref{alg:sampling}
\item Update labels:
\begin{itemize}[leftmargin=0.25cm]
\item Update bias-estimates $\yh(B)$ // Fig.~\ref{alg:bias-estimate}
\item For each $x \in X$ and level $\ell \geq 0$ for which all $\{\yh(B): B \in \B_\ell(x)\}$ are available:
\begin{align}
  \PL_\ell(x) &= \{s \in \{-1, +1\}: \\
             & \mbox{there exists $B \in \B_{\leq \ell}(x)$ s.t. $\yh(B) = s$ and} \notag \\ 
             & \mbox{no $B' \in \B_{\leq \ell}(x)$ has $X_{B'} \subsetneq X_B$.} \} \label{eq:PL} \\ 
\yh_\ell(x) &= 
\left\{
\begin{array}{cl}
+1 & \mbox{if $\PL_\ell(x) = \{+1\}$} \\
-1 &  \mbox{if $\PL_\ell(x) = \{-1\}$} \\
0 & \mbox{if $\PL_\ell(x) = \{\}$} \\
! & \mbox{if $\PL_\ell(x) = \{-1,+1\}$}
\end{array}
\right.
\label{eq:provisional-label1}
\end{align}

@@

In Fig.~\ref{alg:sampling}, we see that the points queried in a ball $B$ at level $\ell$ are $\{z \in X_B: T_z \leq \tau_\ell\}$, which from the definition of $\tau_\ell$ has size $O(k)$. We use these labels to estimate the bias of $B$. Since we need to detect biases of magnitude greater than $\gamma$, we set $k \propto 1/\gamma^2$. 
%This intuition is borne out by the following result, proved in the appendix.

\pagebreak
In the next definition and lemma we establish that the bias estimates for all balls in $\B$ are sufficiently accurate. 
\begin{defn}
\label{def:accurate-bias-estimate}
For any $B \in \B$, bias estimate $\yh(B) \in \{+1,-1,0\}$ is \emph{$\gamma$-accurate} if:
\begin{itemize}
\item $\yh(B) = +1 \implies \eta_X(B) > 0$
\item $\yh(B) = -1 \implies \eta_X(B) < 0$
\item $\yh(B) = 0 \implies |\eta_X(B)| < \gamma$
\end{itemize}
\end{defn}

\begin{lemma}
\label{lemma:accurate-bias-estimate}
  Suppose that $k \geq (192/\gamma^2) \ln (4 |\B|/\delta)$.  Let
  $\yh(B)$ be defined as in Fig.~\ref{alg:bias-estimate}.  Then with
  probability $\geq 1-\delta$, all the $\yh(B)$, for $B \in \B$, are
  $\gamma$-accurate in the sense of
  Definition~\ref{def:accurate-bias-estimate}.
\end{lemma}
Henceforth assume $\yh(B)$ is $\gamma$-accurate for all  $B \in \B$.

@@

Recall that $\PL_\ell(x) \subset \{-1,+1\}$ denotes the possible labels for $x$ given information from balls at levels $\leq \ell$, and is used to assign a value $\yh_\ell(x) \in \{-1,+1,0,!\}$. Despite randomness in querying, it is possible to define two critical levels for each $x$: a level $L_1(x)$ by which $\PL_\ell(x)$ reliably contains the correct label of $x$, and a level $L_2(x)$ by which $\PL_\ell(x)$ reliably omits the wrong label. Intuitively, background sampling dominates until level $L_1(x)$ is eached while focused sampling dominates between $L_1(x)$ and convergence at $L_2(x)$.
Therefor, roughly speaking, the query complexity for point $x$ is low when $L_2(x)-L_1(x)$ is large relative to $L_1(x)$. The precise statement is given in Theorem~\ref{thm:queryComplexity}.

@@

\begin{thm}[Local query complexity] \label{thm:queryComplexity}
Take $k \geq (192/\gamma^2) \ln (4 |\B|/\delta)$. With probability at least $1-2\delta$, the following holds simultaneously for all $x \in X$. Let $L_1(x)$ and $L_2(x)$ be the critical levels for $x$, as in Definition~\ref{defn:L12}. If $L_2(x) \leq \lg (n/2k)$, let
\[
m_o(x) = 32k \cdot \max\bigg( 2^{L_1(x)}, \ \frac{1}{n} \sum_{\ell=L_1(x)+1}^{L_2(x)} |\Delta_\ell| \, 2^\ell\bigg) .
\]
Once the active learner has made $m_o(x)$ queries, $\yh(x)$ is fixed at the Bayes-optimal label $g^*(x)$.
\label{thm:label-complexity}
\end{thm}
Let's interpret this bound. For a specific $x \in X$, there is no telling how $\yh_\ell(x)$ will behave at levels $\ell < L_1(x)$. But for $\ell \geq L_1(x)$, either $\yh_\ell(x) = \, !$, in which case $x$ is in the uncertainty region, or the label of $x$ is correctly set once and for all, which occurs by level $L_2(x)$.

@@

\begin{figure}[t]
\begin{center}
  \includegraphics[width=2.5in,height=1.7in]{figures/Monotone.png}
%\hspace{.5in}
\includegraphics[width=2.5in,height=1.7in]{figures/Massart.png}
\end{center}
\caption{(a) The conditional probability function $\eta$ is monotonically increasing; $\eta(x) = -\gamma, 0, \gamma$ at $x = \lambda_L, \lambda, \lambda_R$, respectively. (b) On each subinterval, $\eta$ is either $> \gamma$ or $<-\gamma$.}
\label{fig:oned}
\end{figure}

@@

\subsection{Example: Massart noise}
\label{sec:example-massart}

In Section~\ref{sec:oned-massart}, we develop another one-dimensional example, in which $\X$ consists of several pieces, such that on each piece $\eta$ is either entirely above $\gamma$ or entirely below $-\gamma$ (Fig.~\ref{fig:oned}(b)). Once again, a logarithmic label complexity is obtained. (Details of the proof are in Section~(\ref{sec:oned-massart})

@@

We describe an experiment designed to evaluate our algorithm.
We use a simple 1D problem, described in Section~\ref{sec:example-massart}.
Although this problem is simple, it is outside the scope of most non-parametric active learning algorithms, which assume smoothness of the conditional probability.

We compare two algorithms, the "active" agorithm described here, and the "passive" algorithm which uses only "background queries". We denote the difference between the error of the generated classifier and that of the Bayes optimal classifier by $\epsilon$. From the theory, we expect the query complexity of the passive algorithm to be $O(\frac{1}{\epsilon})$ and of the active algorithm to be $O(\log (\frac{1}{\epsilon})$.

@@

\subsubsection{Accuracy of bias estimates: Proof of Lemma~\ref{lemma:accurate-bias-estimate}}

@@

For part (a), note that some $B_o \in \B_{L_1(x)}(x)$ has significant bias (that is, bias $\geq \gamma$) towards the correct label $s(x)$, as does any ball $B \in \B(x)$ contained within it. For $\ell \geq L_1(x)$, the set $\{B \in \B_{\leq \ell}(x): X_{B} \subseteq X_{B_o}\}$ is nonempty (it contains $B_o$); pick any minimal ball $B$ within it. Then $\yh(B) = s(x)$ by the $\gamma$-accuracy of bias estimates (Lemma~\ref{lemma:accurate-bias-estimate}) and thus $s(x) \in \PL_\ell(x)$.

@@

\begin{figure}[t]
\begin{center}
\includegraphics[width=3in]{figures/curvature.pdf}
\end{center}
\caption{{\bf A 2D example with bounded curvature:} {\em Left:} We have $\eta = -0.2$ for the orange/red part of the domain and $\eta=0.2$ for the light and dark blue. The orange and light blue indicate the boundary sets for negative and positive respectively. {\em Right:} examples of the $L_1,L_2$ balls for different points. The green $L_1$ circles are the largest monochrome circle that contains the point. The purple $L_2$ circles are the union of all circles of half the radius or smaller. All of these circles are monochrome.}
\label{fig:curvature}
\end{figure}
